Spending too much time at the Galactic bar: chaotic fanning of the Ophiuchus stream

### 2016-01-22 ms.tex
brani and kathryn comments and abstract

\title{Spending too much time at the Galactic bar: Chaotic fanning of the Ophiuchus stream}
\author{Adrian M. Price-Whelan\altaffilmark{\colum,\adrn},
            Branimir Sesar\altaffilmark{\mpia}, 
            Kathryn V. Johnston\altaffilmark{\colum},
            Hans-Walter Rix\altaffilmark{\mpia}
}

@@

\begin{abstract}

% Context
The Ophiuchus stream was recently discovered as an over-density of main sequence stars in Pan-STARRS (PS1) photometry and is peculiar in it's short apparent length given the age of its stellar population.
The stream was later detected in Blue Horizontal Branch (BHB) stars using line-of-sight velocities: Twenty BHB stars in the vicinity of the stream are kinematically distinct from the background halo stars and have distances consistent with the previously detected segment from PS1.
Four of these BHB stars beyond the apparent end of PS1 stars have similarly distinct velocities but have much larger velocity and spatial dispersions. % compared to the PS1 stream segment.
In recent work, we have shown that streams formed on chaotic orbits may display similar peculiar morphologies: Short segments of coherent stellar debris flanked by high-dispersion, low-density ``stream-fans'' at the ends of the stream.
% Aims
The stream's proximity to the triaxial, time-dependent potential of the Galactic bar and the peculiar kinematics and morphology of the stream suggest that it may have formed on a chaotic orbit.
Here we study whether the presence of a triaxial, time-dependent bar can lead to chaotic orbits near (in phase-space) to the Ophiuchus stream. 
% Methods
We fit orbits to the observed stream stars in on static and several barred models for the gravitational potential of the Milky Way. 
We generate model stellar streams along samples from the orbit distributions and compare the maximum-likelihood model streams with the observed density and kinematics of the stream stars.
% Results
With the presence of a bar, many orbits consistent with the stream are strongly chaotic and produce model streams that qualitatively reproduce the apparent shortness and nearby low-density, high-velocity-dispersion stream stars.
% Conclusions
These results motivate efforts to obtain deep photometry of the region surrounding the Ophiuchus stream to detect the predicted low-surface-brightness fans of stellar debris and help distinguish between formation scenarios for the stream.
The existence of or lack of stream-fans may provide an interesting constraint on the triaxiality and pattern speed of the bar.

\end{abstract}

@@

There are a number of peculiarities about the observed kinematics of probable member stars of the Ophiuchus stream. For example, the de-projected length of the visible part of the stream is short given the age of its stellar population ($\approx$1.5 kpc). S15 fit an orbit to the kinematics of the stream stars in a static, axisymmetric model for the gravitational field of the inner Galaxy and ran N-body simulations of globular clusters on this orbit. S15 find that---on this orbit---the portion of the stream visible as an over-density in main-sequence stars must have been formed in the last $\lesssim$400 Myr for the stream to remain as short as it is observed. This dynamical age is at odds with the old stellar population: The stars appear to have formed $\approx$12 Gyr ago, but the cluster apparently fully disrupted at once in the last 400 Myrs. Another puzzle is the existence of blue horizontal branch (BHB) stars close to the stream (within a few degrees) with similar radial velocities, but with a large dispersion in both sky position and velocity \citep[][hereafter S16]{sesar16}. The stream has a very distinct and large line-of-sight velocity ($\approx$290 \kms) and is therefore easily detected above the background halo population. Four BHB stars have been detected with line-of-sight velocities $>230~\kms$ that lie close to an extrapolation of the stream on the sky, but have a velocity dispersion $\approx$75 times larger than the measured internal velocity dispersion of the stream stars. The velocities of these stars are highly inconsistent with belonging to the background halo population (see Section 4, S16). These stars hint at the existence of associated low-density, high-dispersion features that were not modeled in S15 and are not predicted by the $N$-body simulations from this prior work.

@@

The orbit fit and $N$-body simulations in S15 used a static, axisymmetric potential to represent the Milky Way potential, but it is well-known that the Galactic bulge contains a triaxial, bar-like structure several kpc in size \citep[e.g.,][]{blitz91, weinberg92, dwek95, wegg13}. Given the proximity of the stream to the center of the Galaxy, the time-dependent, triaxial potential of the Galactic bar must be taken into account when modeling the orbit of the Ophiuchus stream. The presence of a bar-like perturbation to the potential will change the orbit of the stream progenitor and the orbit structure in the inner galaxy \citep{zotos12, portail15b, gajda15}. Bar-like features can also introduce a significant number of chaotic orbits in their vicinity \citep{weinberg15}.

Recent work has shown that chaos can dramatically alter the density evolution of tidal streams \citep[e.g.,][]{fardal14, apw15-chaos}. Along certain chaotic orbits, the stream stars will spread much faster in 3D position than from ordinary phase-mixing and, depending on the orbital phase at which the stream is observed, may develop large, low-density ``fans'' of stars at the ends of a stream \citep{pearson15, apw15-chaos}. As a first application of this theoretical understanding, we study whether stream-fanning---chaotic or simply from density evolution in a triaxial, time-dependent potential---could plausibly explain the observed properties of the Ophiuchus stream. In particular, we consider whether such models can reproduce:
\begin{enumerate}
	\item the apparent shortness and fast density truncation of the stream;
	\item the increased positional dispersion of the four new candidate members from S16;
	\item the large velocity dispersion of the S16 stars.
\end{enumerate}
Our goal is not to perfectly represent the observed data but rather to explore the plausibility of explaining the peculiarities of the stream using chaotic stream fanning.

In Section~\ref{sec:method} we describe the methods used in this work: in Section~\ref{sec:potential} we describe the models we use for the gravitational potential of the Galaxy, in Section~\ref{sec:mocks} we explain the simple method we use to generate mock streams, and in Section~\ref{sec:chaos-indicator} we briefly describe the methods we use for detecting and characterizing chaotic orbits. In Section~\ref{sec:results1} we discuss the differences in the orbit structure between a static, axisymmetric potential model and a model with a time-dependent bar potential, and in Section~\ref{sec:results2} we use mock streams to argue that [chaotic stream-fanning is a plausible explanation for the observational peculiarities of the Ophiuchus stream.] We conclude in Section~\ref{sec:conclusions}.

@@

Our goal is to (1) assess whether the Galactic bar can produce chaotic orbits in the vicinity of the Ophiuchus stream and (2) determine if chaotic density evolution of tidal debris stripped from the progenitor of this stream can explain the apparent shortness of the stream and low-density, high-dispersion stars beyond the extent of main sequence stars observed in PS1. In this section, we describe the potential models we use to represent the galaxy and describe the methods we use to detect and quantify the strength of chaos for orbits. We then describe the method we use for fitting orbits to the stream stars. Finally, we describe the method we use to generate mock stellar streams.

In this work, we assume the Sun is at Galactocentric position $(x,y,z) = (-8.3,0,0)~{\rm kpc}$ \citep[e.g.,][]{schoenrich12} with velocity $(v_x,v_y,v_z) = (-11.1, 250, 7.25)~\kms$ \citep[e.g.,][]{schoenrich10, schoenrich12}.

@@

The BFE representation fixes the axis ratios of the bar---that is, the the exponential scale lengths along the three axes of the bar were adopted from \cite{dwek95} when the expansion coefficients were calculated in W12); all other potential parameter values are given in Table~\ref{tbl:potential-params-barred}. The mass of the halo is fixed and the mass of the disk and bar are varied in order to qualitatively reproduce the flatness and amplitude of the circular velocity curve of the Milky Way \citep{bovy12}. Figure~\ref{fig:potentials}, top left shows the circular velocity along the line connecting the Sun to the Galactic center in this model (the Galactic $x$ axis). Figure~\ref{fig:potentials}, top right shows contours of constant surface density for a face-on (left) and edge-on (right) view of this potential model with the bar angle set to $20^\circ$ \citep[compare to, e.g., Figure 3 in][]{portail15}. We consider a grid of nine parameter combinations of bar angle and pattern speed. These model names and parameter values are given in Table~\ref{tbl:bar-specific}.

@@

\subsection{Fitting orbits to the Ophiuchus stream}

@@

To generate mock stellar streams, we use the method presented in \citep{fardal14}: star particles are `released' from a progenitor system near the Lagrange points with a dispersion in position and velocity that is set by the mass and orbit of the progenitor. We draw samples from the posterior probability distributions over orbital parameters from fitting orbits to the stream star members and use these as the progenitor orbital parameters (Section~\ref{sec:orbitfit}). For a given progenitor orbit---the 6D position of the orbit today---we integrate the orbit backwards in time for a given integration period. From the endpoint of the backwards-integration (e.g., the past position), we begin integrating the orbit forward in time, but now at each time-step a star particle is released near each of the Lagrange points of the progenitor. The position of the Lagrange points and the scale of the dispersion in position and velocity are set by the progenitor mass, $m$. The star particles are drawn from Gaussians centered on the Lagrange points (in position) and the progenitor (in velocity) and the full parametrization of the release distribution is given in \cite{fardal14}. This method has been shown to reproduce the morphologies of $N$-body simulations of stellar streams, but requires far less computing time because it relies only on integrating test-particle orbits.

@@

Figure~\ref{fig:orbitfits} summarizes our results from fitting orbits to the BHB stream stars. Shown in each panel are the Ophiuchus stream star members (black points) and orbits integrated from samples from the posterior probability over orbital parameters (blue lines). We only show one of the barred potentials: The end-to-end integration time of the orbit over the observed extent of the stream is only $\approx$6 Myr, so the derived orbits are extremely similar in all potentials (the time-dependence of the bar potential is not significant over such short timescales). The orbital periods are typically $\approx$170--200 Myr with pericenters $r_p \approx 4~{\rm kpc}$ and apocenters $r_a \approx 12$--$15~{\rm kpc}$. Though the coordinate and velocity parameter values in observed coordinates are very similar between each potential model, the resulting orbits are quite different. For the posterior samples in each potential, we take the mean values of the coordinate and velocity parameters and convert to Galactocentric coordinates (e.g., Table~\ref{tbl:param-means}). Figure~\ref{fig:orbits-yz} shows projections of these ``mean'' orbits in each potential model.

@@

% ---------------------------------------------------------------------------------
\begin{figure}[!tbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/orbit-yz}
\caption{  Projections of orbits integrated from the mean orbital parameters estimated from the orbit fitting posterior distributions in each potential model. Orbits are integrated for 6 Gyr and shown in Galactocentric, Cartesian coordinates. Even though the mean orbital parameters have nearly identical values (e.g., the initial conditions are nearly identical in Heliocentric coordinates), the orbits in each potential model are  different in appearance. }
\label{fig:orbits-yz}
\end{center}
\end{figure}

@@

% ---------------------------------------------------------------------------------
\begin{figure}[!tbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/lyapunov-hist}
\caption{ Histograms of estimates of the Lyapunov time, $\lyapt=1/\lyapexp$, for posterior samples from the orbit fitting procedure (Section~\ref{sec:orbitfit}). Vertical lines show the Lyapunov time value of the mean orbit in that potential model. Orbits are typically more chaotic (have shorter Lyapunov times) for higher pattern speeds. There is little dependence on bar angle.}
\label{fig:lyapunov-hist}
\end{center}
\end{figure}

\section{Results 2: Stream models for the Ophiuchus stream}

@@

We randomly sample 256 orbits from the orbital parameter posterior distributions and generate mock streams along each orbit. We use the method outlined above (Section \ref{sec:mocks}) to generate the streams and set the free parameters as follows: (1) we evolve the progenitors for 6 Gyr along each orbit to explore stream models where the shortness of the stream is \emph{not} due to an instantaneous disruption 400 Myrs ago, (2) we release star particles every 0.5 Myr (uniformly in time) to densely sample the final density distribution, (3) we set the progenitor mass to $m=10^4~\msun$ (as was estimated by S15), and (4) we set the disruption time of the progenitor equal to the last time at which a pericenter and disk crossing coincide (in each case, this is at $t \approx -200~{\rm Myr}$). After the disruption time, we continue releasing the star particles uniformly in time rather than releasing a ``burst'' of particles at once. We therefore expect that the density of the most recently disrupted debris will be systematically higher for the model streams as compared to the observed stream.

@@

	\begin{tabular}{ccc}
	\toprule
%	name & $\phi_2$ [deg] & $d$ [kpc] & $\mu_l$ [mas yr$^{-1}$] & $\mu_b$ [mas yr$^{-1}$] & $v_r$ [km s$^{-1}$] & $s_{\phi_2}$ [deg] & $s_{d}$ [kpc] & $s_{v_r}$ [km s$^{-1}$]\\\midrule
%static & $-0.03\pm0.05$ & $8.35\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.1\pm0.9$ & $0.20\pm0.04$ & $0.31\pm0.10$ & $2.9\pm0.8$\\
%bar1--9 & $-0.03\pm0.05$ & $8.40\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.0\pm1.0$ & $0.21\pm0.05$ & $0.31\pm0.14$ & $3.2\pm0.9$\\
	$s_{\phi_2}$ [deg] & $s_{d}$ [kpc] & $s_{v_r}$ [km s$^{-1}$]\\\midrule
	$0.20\pm0.04$ & $0.31\pm0.10$ & $2.9\pm0.8$\\
	$0.21\pm0.05$ & $0.31\pm0.14$ & $3.2\pm0.9$\\
	\bottomrule
	\end{tabular}
	\caption{Estimated mean and standard deviation of samples from the marginal posterior distributions over each parameter in our orbit fit model. For the barred potentials, all mean values are the same because the time-dependence of the bar doesn't impact the orbit fit over the short length of the stream. We have made samples from the full posterior distribution available with this article and provide code to transform to and from stream coordinates (see http://adrian.pw/ophiuchus/coordinates for more information).\label{tbl:param-means} }
\end{center}
\end{table*}

For each generated mock stream, we compute the likelihood of the data (now including all BHB stars from S15 and S16, e.g., all points in Figure~\ref{fig:orbitfits}) given the star particles by estimating the phase-space model density using a kernel density estimate with a Gaussian kernel \citep[see, e.g.,][]{bonaca14}. For each $i$ data point $\bs{x}_i$ and each $k$ model point with coordinates $\bs{y}_k$, we compute the likelihood by converting the model point position and velocity into Heliocentric coordinates and evaluate
\begin{align}
	p(\bs{x}_i \given \bs{y}_k, \bs{\sigma}_i, \bs{h}) = \mathcal{N}(\bs{x}_i \given \bs{y}_k, \bs{\sigma}_i^2 +\bs{h}^2)
\end{align}
where $\mathcal{N}(x\given \mu,\sigma^2)$ is the normal distribution with mean $\mu$ and variance $\sigma^2$ and $\bs{h}$ represents the diagonal of the bandwidth matrix, $\bs{\rm H}$, used for the density estimate, $\bs{h} = {\rm diag}(\bs{\rm H})$. We fix the bandwidth parameters as follows
\begin{equation}
	\bs{h} = \left(
	\begin{array}{c}
	0.2~{\rm deg}\\
	0.2~{\rm deg}\\
	0.2~{\rm kpc}\\
	0.2~\kms
	\end{array}
	\right)
\end{equation} %(max$\mathcal{L}$)
for sky position in Galactic coordinates ($l$, $b$), distance, and radial velocity. The full likelihood for all $N$ data points given $K$ model points is 
\begin{equation}
	\mathcal{L} = \prod_i^N \frac{1}{K} \sum_k^K p(\bs{x}_i \given \bs{y}_k, \bs{\sigma}_i, \bs{h}).
\end{equation}

>>>>

\title{Spending too much time at the Galactic bar: chaotic fanning of the Ophiuchus stream}
\author{Adrian M. Price-Whelan\altaffilmark{\colum,\adrn},
            Branimir Sesar\altaffilmark{\mpia}, 
            Kathryn V. Johnston\altaffilmark{\colum},
            Hans-Walter Rix\altaffilmark{\mpia}
}

@@

\begin{abstract}
The Ophiuchus stellar stream is peculiar: (1) its length is short given the age of its constituent stars, and (2) several probable member stars that lie close in both sky position and velocity have dispersions in these dimensions that far exceed those seen within the stream. 
The stream's proximity to the Galactic center suggests that the bar should have a significant influence on its dynamical history: The triaxiality and time-dependence of the bar may generate chaotic orbits in the vicinity of the stream that can greatly affect its morphology. 
We explore this hypothesis with models of stream formation along orbits consistent with Ophiuchus' properties in a Milky Way potential model that includes a rotating bar.
We find that in all choices for the rotation parameters of the bar, orbits fit to the stream are strongly chaotic.
Mock streams generated along these orbits qualitatively match the observed properties of the stream: because of chaos, stars stripped early generally form low-density, high-dispersion ``fans'' leaving only the most recently disrupted material detectable as a strong over-density. 
These results motivate efforts to obtain deep photometry and spectroscopy of the region surrounding the stream to detect this predicted low-surface-brightness tidal debris.
The existence of or lack of these features would provide an interesting constraint on the properties of the Milky Way bar and would help distinguish between formation scenarios for the stream.
This is the first time that chaos has been used to explain the properties of a stellar stream and is the first demonstration of the dynamical importance of chaos in the Galactic halo.
\end{abstract}

@@

There are a number of peculiarities about the observed kinematics of the Ophiuchus stream. For example, the de-projected length of the visible part of the stream is short given the age of its stellar population ($\approx$1.5 kpc). S15 fit an orbit to the kinematics of the stream stars in a static, axisymmetric model for the gravitational field of the inner Galaxy and ran N-body simulations of globular clusters on this orbit. S15 find that---on this orbit---the portion of the stream visible as an over-density in main-sequence stars must have been formed in the last $\lesssim$400 Myr for the stream to remain as short as it is observed. This dynamical age is at odds with the old stellar population: The stars appear to have formed $\approx$12 Gyr ago, but the abrupt end of the stream suggests that the cluster apparently fully disrupted at once in the last 400 Myrs. Another puzzle is the existence of blue horizontal branch (BHB) stars close to the stream (within a few degrees) with similar radial velocities, but with a large dispersion in both sky position and velocity \citep[][hereafter S16]{sesar16}. The stream has a very distinct and large line-of-sight velocity ($\approx$290 \kms) and is therefore easily detected above the background halo population. Four BHB stars have been detected with line-of-sight velocities $>230~\kms$ that lie close to an extrapolation of the stream on the sky, but have a velocity dispersion $\approx$75 times larger than the measured internal velocity dispersion of the stream stars. The velocities of these stars are highly inconsistent with belonging to the background halo population (see Section 4, S16). These stars hint at the existence of associated low-density, high-dispersion features that were not modeled in S15 and are not predicted by the $N$-body simulations from this prior work, which assume a sudden, total disruption of the stream progenitor.

@@

The orbit fit and $N$-body simulations in S15 used a static, axisymmetric potential to represent the Milky Way potential, but it is well-known that the Galactic bulge contains a triaxial, rotating, bar-like structure several kpc in size \citep[e.g.,][]{blitz91, weinberg92, dwek95, wegg13}. Given the proximity of the stream to the center of the Galaxy, the time-dependent, triaxial potential of the Galactic bar must be taken into account when modeling the orbit of the Ophiuchus stream. The presence of a bar-like perturbation to the potential will change the orbit of the stream progenitor and the orbit structure in the inner galaxy \citep{zotos12, portail15b, gajda15}. Bar-like features can also introduce a significant number of chaotic orbits in their vicinity \citep{weinberg15} and generate resonances that may also affect stream formation \citep{hattori15}.

Recent work has shown that chaos can dramatically alter the density evolution of tidal streams \citep[e.g.,][]{fardal14, apw15-chaos}. Along certain chaotic orbits, the stream stars will spread much faster in 3D position than from ordinary phase-mixing and, depending on the orbital phase at which the stream is observed, may develop large, low-density ``fans'' of stars at the ends of a stream \citep{pearson15, apw15-chaos}. As a first application of this theoretical understanding, we study whether stream-fanning---chaotic or simply from density evolution in a triaxial, time-dependent potential---could plausibly explain the observed properties of the Ophiuchus stream. In particular, we consider whether such models can reproduce:
\begin{enumerate}
	\item the apparent shortness and fast density truncation of the stream;
	\item the increased positional dispersion of the four new candidate members from S16;
	\item the large velocity dispersion of the S16 stars.
\end{enumerate}
Our goal is not to perfectly represent the observed data but rather to explore the plausibility of explaining the peculiarities of the stream using chaotic stream fanning. Note that an alternate model was recently proposed that instead places the stream progenitor on an orbit in resonance with the bar. We discuss the differences between these two models in Section~\ref{sec:discussion}.

In Section~\ref{sec:method} we describe the methods used in this work: in Section~\ref{sec:potential} we describe the models we use for the gravitational potential of the Galaxy, in Section~\ref{sec:orbitfit-nonapdx} we outline the probabilistic procedure we use to fit orbits to the stream data (explained in detail in Appendix~\ref{sec:orbitfit}), and in Section~\ref{sec:mocks} we explain the simple method we use to generate mock streams. In Section~\ref{sec:results1} we discuss the results from fitting orbits to the data in a static, axisymmetric potential model and several potential models with a time-dependent bar. In Section~\ref{sec:results2} we generate mock streams along these orbits to argue that chaotic stream-fanning is a plausible explanation for the observational peculiarities of the Ophiuchus stream. We discuss the implications of this work and possibilities for future work in Section~\ref{sec:discussion} and we conclude in Section~\ref{sec:conclusions}.

@@

Our goal is to (1) assess whether the Galactic bar can produce chaotic orbits in the vicinity of the Ophiuchus stream and (2) determine if chaotic density evolution of tidal debris stripped from the progenitor of this stream can explain the apparent shortness of the stream and low-density, high-dispersion stars beyond the extent of main sequence stars observed in PS1. In this section, we describe the potential models we use to represent the galaxy and outline the methods we use to detect and quantify the strength of chaos for individual orbits. We then describe the likelihood function we use for fitting orbits to the stream stars. Finally, we describe how we generate mock stellar streams for a progenitor on a given orbit.

Throughout we assume the Sun is at Galactocentric position $(x,y,z) = (-8.3,0,0)~{\rm kpc}$ \citep[e.g.,][]{schoenrich12} with velocity $(v_x,v_y,v_z) = (-11.1, 250, 7.25)~\kms$ \citep[e.g.,][]{schoenrich10, schoenrich12}.

@@

The BFE representation fixes the axis ratios of the bar---that is, the exponential scale lengths along the three axes of the bar were adopted from \cite{dwek95} when the expansion coefficients were calculated in W12; all other potential parameter values are given in Table~\ref{tbl:potential-params-barred}. The mass of the halo is fixed and the mass of the disk and bar are varied in order to qualitatively reproduce the flatness and amplitude of the circular velocity curve of the Milky Way \citep{bovy12}. Figure~\ref{fig:potentials}, top left shows the circular velocity along the line connecting the Sun to the Galactic center in this model (the Galactic $x$ axis). Figure~\ref{fig:potentials}, top right shows contours of constant surface density for a face-on (left) and edge-on (right) view of this potential model with the bar angle set to $20^\circ$ \citep[compare to, e.g., Figure 3 in][]{portail15}. We consider a grid of nine parameter combinations of bar angle and pattern speed. Model names and parameter values are given in Table~\ref{tbl:bar-specific}.

@@

\subsection{Fitting orbits to the Ophiuchus stream}\label{sec:orbitfit-nonapdx}

@@

To generate mock stellar streams, we use the method presented in \citet{fardal14}: Star particles are `released' from a progenitor system near the Lagrange points with a dispersion in position and velocity that is set by the mass and orbit of the progenitor. We draw samples from the posterior probability distributions over orbital parameters from fitting orbits to the stream star members and use these as the progenitor orbital parameters (Section~\ref{sec:orbitfit}). For a given progenitor orbit---the 6D position of the orbit today---we integrate the orbit backwards in time for a given integration period. From the endpoint of the backwards-integration (e.g., the past position), we begin integrating the orbit forward in time, but now at each time-step a star particle is released near each of the Lagrange points of the progenitor. The position of the Lagrange points and the scale of the dispersion in position and velocity are set by the progenitor mass, $m$. The star particles are drawn from Gaussians centered on the Lagrange points (in position) and the progenitor (in velocity) and the full parametrization of the release distribution is given in \cite{fardal14}. This method has been shown to reproduce the morphologies of $N$-body simulations of stellar streams, but requires far less computing time because it relies only on integrating test-particle orbits.

@@

Figure~\ref{fig:orbitfits} summarizes our results from fitting orbits to the BHB stream stars. Shown in each panel are the high-probability Ophiuchus stream stars (black points, to which orbits are fit) and orbits integrated from samples from the posterior probability over orbital parameters (blue lines). The four ``fanned'' BHB stars from S16 are shown as grey squares and are not included in the orbit fitting procedure. We only show one of the barred potentials: The end-to-end integration time of the orbit over the observed extent of the stream is only $\approx$6 Myr, so the derived orbits are extremely similar in all potentials (the time-dependence of the bar potential is not significant over such short timescales). The orbital periods are typically $\approx$170--200 Myr with pericenters $r_p \approx 4~{\rm kpc}$ and apocenters $r_a \approx 12$--$15~{\rm kpc}$. Though the coordinate and velocity parameter values in observed coordinates are very similar between each potential model, the resulting orbits are quite different. For the posterior samples in each potential, we take the mean values of the coordinate and velocity parameters and convert to Galactocentric coordinates (e.g., Table~\ref{tbl:param-means}). Figure~\ref{fig:orbits-yz} shows projections of these ``mean'' orbits in each potential model.

@@

% ---------------------------------------------------------------------------------
\begin{figure}[!tbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/orbit-yz}
\caption{  Projections of orbits integrated from the mean orbital parameters estimated from the orbit fitting posterior distributions in each potential model. Orbits are integrated for 6 Gyr and shown in Galactocentric, Cartesian coordinates. Even though the mean orbital parameters have nearly identical values (e.g., the initial conditions are nearly identical in heliocentric coordinates), the orbits in each potential model are  different in appearance. }
\label{fig:orbits-yz}
\end{center}
\end{figure}

@@

% ---------------------------------------------------------------------------------
\begin{figure}[!tbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/lyapunov-hist}
\caption{ Histograms of estimates of the Lyapunov time, $\lyapt=1/\lyapexp$, for posterior samples from the orbit fitting procedure (Section~\ref{sec:orbitfit}). Orbits are typically more chaotic (have shorter Lyapunov times) for higher pattern speeds. There is little dependence on bar angle.}
\label{fig:lyapunov-hist}
\end{center}
\end{figure}

\section{Results 2: Stream models for the Ophiuchus stream}\label{sec:results2}

@@

For each potential model, we randomly sample 256 orbits from the orbital parameter posterior distributions and generate mock streams along each orbit. We use the method outlined above (Section \ref{sec:mocks}) to generate the streams and set the free parameters as follows: (1) we evolve the progenitors for 6 Gyr along each orbit prior to the current position to explore stream models where the shortness of the stream is \emph{not} due to an instantaneous disruption 400 Myrs ago, (2) we release star particles every 0.5 Myr (uniformly in time) to densely sample the final density distribution, (3) we set the progenitor mass to $m=10^4~\msun$ (as was estimated by S15), and (4) we set the disruption time of the progenitor equal to the last time at which a pericenter and disk crossing coincide (in each case, this is at $t \approx -200~{\rm Myr}$). After the disruption time, we continue releasing the star particles uniformly in time (every 0.5 Myr) rather than releasing a ``burst'' of particles at once. We therefore expect that the density of the most recently disrupted debris will be systematically higher for the model streams as compared to the observed stream.

@@

	\begin{tabular}{ccc}
	\toprule
%	name & $\phi_2$ [deg] & $d$ [kpc] & $\mu_l$ [mas yr$^{-1}$] & $\mu_b$ [mas yr$^{-1}$] & $v_r$ [km s$^{-1}$] & $s_{\phi_2}$ [deg] & $s_{d}$ [kpc] & $s_{v_r}$ [km s$^{-1}$]\\\midrule
%static & $-0.03\pm0.05$ & $8.35\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.1\pm0.9$ & $0.20\pm0.04$ & $0.31\pm0.10$ & $2.9\pm0.8$\\
%bar1--9 & $-0.03\pm0.05$ & $8.40\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.0\pm1.0$ & $0.21\pm0.05$ & $0.31\pm0.14$ & $3.2\pm0.9$\\
	$s_{\phi_2}$ [deg] & $s_{d}$ [kpc] & $s_{v_r}$ [km s$^{-1}$]\\\midrule
	$0.20\pm0.04$ & $0.31\pm0.10$ & $2.9\pm0.8$\\
	$0.21\pm0.05$ & $0.31\pm0.14$ & $3.2\pm0.9$\\
	\bottomrule
	\end{tabular}
	\caption{Estimated mean and standard deviation of samples from the marginal posterior distributions over each parameter in our orbit fit model (the posterior distributions are very close to Gaussian). For the barred potentials, all mean values are the same because the time-dependence of the bar doesn't impact the orbit fit over the short length of the stream. We have made samples from the full posterior distribution available with this article and provide code to transform to and from stream coordinates (see http://adrian.pw/ophiuchus/coordinates for more information).\label{tbl:param-means} }
\end{center}
\end{table*}

For each generated mock stream, we compute the likelihood of the data (now including all BHB stars from S15 and S16, e.g., all points in Figure~\ref{fig:orbitfits}) given the star particles by estimating the phase-space model density using a kernel density estimate with a Gaussian kernel \citep[see, e.g.,][]{bonaca14}. For each $i$ data point $\bs{x}_i$ and each $k$ model point with coordinates $\bs{y}_k$, we compute the likelihood by converting the model point position and velocity into heliocentric coordinates and evaluate
\begin{align}
	p(\bs{x}_i \given \bs{y}_k, \bs{\sigma}_i, \bs{h}) = \mathcal{N}(\bs{x}_i \given \bs{y}_k, \bs{\sigma}_i^2 +\bs{h}^2)
\end{align}
where $\mathcal{N}(x\given \mu,\sigma^2)$ is the normal distribution with mean $\mu$ and variance $\sigma^2$ and $\bs{h}$ represents the diagonal of the bandwidth matrix, $\bs{\rm H}$, used for the density estimate, $\bs{h} = {\rm diag}(\bs{\rm H})$. We fix the bandwidth parameters as follows
\begin{equation}
	\bs{h} = \left(
	\begin{array}{c}
	0.2~{\rm deg}\\
	0.2~{\rm deg}\\
	0.2~{\rm kpc}\\
	0.2~\kms
	\end{array}
	\right)
\end{equation} %(max$\mathcal{L}$)
for sky position in Galactic coordinates ($l$, $b$), distance, and radial velocity. The full likelihood for all $N$ data points given $K$ model points is 
\begin{equation}
	\mathcal{L} = \prod_i^N \frac{1}{K} \sum_k^K p(\bs{x}_i \given \bs{y}_k, \bs{\sigma}_i, \bs{h}).
\end{equation}

Figures~\ref{fig:mockstream0}--\ref{fig:mockstream1} show the final particle positions and line-of-sight velocities in heliocentric coordinates for the maximum likelihood mock streams (grey points) in each potential. Vertical, dashed lines show the approximate extent of the part of the stream visible in main-sequence stars (excluding the BHB stars from S16). There are a few interesting features to note from these panels: 
\begin{enumerate}
	\item Even in the static potential (Figure~\ref{fig:mockstream0}, leftmost column), there is a slight decrease in the density for the model stream towards higher Galactic longitudes, $l$. This is a projection effect: The portion of the stream at larger $l$ is closer points almost directly towards the Sun so that the debris covers a larger area. 
	\item The density of the mock stream in the static potential decreases slowly rather than abruptly as is observed. This is shown in the top panels of each column where each contour level represents a factor of 10 difference in projected surface density. In the static potential model the length of dense debris extends much farther than the observed extent of the stream (vertical dashed lines), whereas in some of the barred potential models the stars released earlier have ``fanned'' and are associated with much lower density debris (e.g., bar8).
	\item The four high-dispersion BHB stars beyond the end of the stream (from S16) don't match in position and velocity with the particle distribution from even the maximum likelihood stream model in the static potential. However, in some barred potentials the chaotic evolution of the stream stars can lead to over-densities of stars with an increased positional dispersion and significantly discrepant velocities (e.g., bar8).
	\item None of the stream models---static or barred---produce an appreciable density of stars with line-of-sight velocities near the S16 BHB star with the largest velocity ($\approx 320~\kms$). This star is either an interesting Ophiuchus member star or is associated with some other kinematic substructure.
	\item For the barred potentials, the stream morphology is very sensitive to the properties of the bar (especially the pattern speed) and to the initial conditions of the orbit. We have found that the morphology can vary significantly between nearby orbits in the same potential model (because these are strongly chaotic orbits), but the overall characteristics remain similar: along more strongly chaotic orbits, the debris ``fans'' more and the apparent dense part of the model streams is shorter.
\end{enumerate}

The viewing angle and stream geometry is more clearly demonstrated in Figure~\ref{fig:mockstreamxyz}, which shows $x$-$z$ projections of the star particles in Galactocentric Cartesian coordinates (grey) along with the position of the Sun (symbol at $(x,z)=(-8.3,0)~{\rm kpc}$) and the ``window'' of the heliocentric, sky-position plots of Figures~\ref{fig:mockstream0}--\ref{fig:mockstream1} (shown as blue lines).

The density truncation of the mock streams in each potential model is more clearly seen in terms of the density contrast between stream stars and background stars, visualized in Figure~\ref{fig:densitymaps}: This figure shows mock sky-density maps of stars generated by superimposing the maximum likelihood model streams over a noisy background of stars. The number of mock stream star particles used to generate the map has been normalized such that the total number of stars within the observed extent of the stream (between $5.85 > l > 3.81~{\rm deg}$) is equal to the number of stars attributed to the Ophiuchus stream in the PS1 data \citep[$N \approx 500$][]{bernard14}.
